---
title: "relatedness.compare"
output: html_document
date: "2024-06-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

load in datata
```{r}
library(data.table)
library(tidyverse)
library(foreach)
library(readxl)
library(gmodels)

setwd("/Users/supad/OneDrive/Documents/Bergland Research/R_data_objects/May_2024_objects/")
ldgrm = readRDS("LDGRM")
fullgrm = readRDS("GRM.Aug")

locogrm2l = readRDS("No2LGRM")
locogrm2r = readRDS("No2RGRM")
locogrm3l = readRDS("No3LGRM")
locogrm3r = readRDS("No3RGRM")
locogrmx = readRDS("NoXGRM")

#inversion status of dgrp lines

```
next we want to find a way to create a long form data table with the relatedness between lines. we want one measure of relatedness between two unique lines- no repeats, and no lines against each other. 

```{r}
#we'll try melting to start
#convert matrix to data tabl
ldgrm = as.data.table(ldgrm)
ldgrm$ids = colnames(ldgrm)
ldmelt = melt(ldgrm, id.vars = c("ids"), variable.name = "id2")
#remove any row where id1 = id2
#and any duplicate combo of id1 and id2
ldmelt$id2 = as.character(ldmelt$id2)
#remove 
ld.r = subset(ldmelt, ids != id2 & 
             !duplicated(cbind(pmin(ids, id2),
                               pmax(ids, id2))))
ld.r$method = "LDpruned"
#now for full grm
fullgrm = as.data.table(fullgrm)
fullgrm$ids = colnames(fullgrm)
fullmelt = melt(fullgrm, id.vars = c("ids"), variable.name = "id2")
#remove any row where id1 = id2
#and any duplicate combo of id1 and id2
fullmelt$id2 = as.character(fullmelt$id2)
#remove 
full.r = subset(fullmelt, ids != id2 & 
             !duplicated(cbind(pmin(ids, id2),
                               pmax(ids, id2))))
full.r$method = "Fullgrm"

```

For loco, we'll make a quick forloop

```{r}
chrom = c("2L", "2R", "3L" ,"3R", "X")
locolist = list(locogrm2l, locogrm2r, locogrm3l, locogrm3r,locogrmx)
out = foreach(f = c(1:5)) %do% {
  #f =1
  #now for loco
  locogrm = as.data.table(locolist[f])

locogrm$ids = colnames(locogrm)
locomelt = melt(locogrm, id.vars = c("ids"), variable.name = "id2")
#remove any row where id1 = id2
#and any duplicate combo of id1 and id2
locomelt$id2 = as.character(locomelt$id2)
#remove 
loco.r = subset(locomelt, ids != id2 & 
             !duplicated(cbind(pmin(ids, id2),
                               pmax(ids, id2))))
loco.r$method = "locogrm"
loco.r$chr = chrom[f]
loco.r
}
loco.out = rbindlist(out)
#now we find the median relatedness per dgrp comparison
# loco.out$combo.id = paste(loco.out$ids, loco.out$id2, sep = "-")
# loco.out = loco.out %>% 
#   group_by(ids, id2, method)%>% 
#   summarise(value = median(value))

#create one for the no grm- same thing but all zeros
no.grm = loco.out
no.grm$value = 0
no.grm$method = "No_GRM"
             
```
now bind and graph the data

```{r}
colnames(no.grm)
colnames(loco.out)
colnames(full.r)
colnames(ld.r)

talldata = rbind( full.r, ld.r)
talldata$chr = "2L"
talldata = rbind(talldata, loco.out, no.grm)
#make a box plot with errorbars
talldata %>% 
  group_by(method) %>% 
   summarise(mean = ci(value)[1],
            uci = ci(value)[2],
            lci = ci(value)[3] ) %>% 
ggplot(
  aes(x = method,
      y = mean,
      ymin = lci,
      ymax = uci)
)+
  geom_bar(stat = "identity") + 
   geom_errorbar(width = 0.1, position=position_dodge(width = 0.5), show.legend = F) 
  
```

now try a histogram like in the huang paper
```{r}
ggplot(talldata, aes(x = value))+ geom_histogram(
  bins = 80
)+
  facet_grid(method~., scales = "free")+
  xlim(-0.075, 0.15)
```
load in the inversion data, and filter to only include pairwise comparison between lines with the same karyotype
```{r}

inv.dt = read_xlsx("inversion.xlsx")
#fix up inv.dt data prior to joining
merge.dt = inv.dt %>% 
  select(c(`DGRP Line`, `In(2L)t`, `In(2R)NS`, `In(3R)P`, `In(3R)K`, `In(3R)Mo`)) %>% 
  mutate(`DGRP Line` = gsub("DGRP", "line", `DGRP Line`)) %>% 
  rename("ids" = "DGRP Line") %>% 
  pivot_longer(
    cols = !ids,
    names_to = "inversion",
    values_to = "status"
  ) %>% 
  as.data.table(.)
mergedata = merge(talldata, merge.dt, by = c("ids"), allow.cartesian = T)
#now we merge them again based on second id
merge.dt2 = merge.dt %>% 
  rename("id2" = "ids")
mergedata = merge(mergedata, merge.dt2, by = "id2", allow.cartesian = T)#note, inverstion status of id"1" is now marked byx, id2 is marked by y

#for now, focus on pairwise combinations where the lines are the same karyotype, split across each inversion
samedata = mergedata %>% 
  group_by(inversion.x) %>% 
  filter(inversion.x == inversion.y) %>% #compare across the same inversion
  filter(status.x == status.y) %>%#filter to the same inversion type
  as.data.table(.)


```
it turns out the above approach- bracketed comparisions to only show flys of the same karyotype- doesn't show anything. lets' try to compare standard(fully un inverted) to comparisons between flies homozgous for the same inversion. for now, these inverted flies can have other inversions present as well. 
```{r}
#we want list of dgrp lines that are homozgyous for standard, and then each inversion.
head(inv.dt)
head(merge.dt)
st.dt = inv.dt %>% 
  select(c(`DGRP Line`, `In(2L)t`, `In(2R)NS`, `In(3R)P`, `In(3R)K`, `In(3R)Mo`)) %>% 
  mutate(`DGRP Line` = gsub("DGRP", "line", `DGRP Line`)) %>% 
  filter( `In(2L)t` == "ST",
          `In(2R)NS` == "ST",
          `In(3R)P` == "ST",
          `In(3R)K` == "ST",
          `In(3R)Mo` == "ST",
  )
st.lines = as.vector(st.dt[[1]])#list of lines with no major inversions
st.data = talldata %>% 
  filter(ids %in% st.lines) %>% 
  filter(id2 %in% st.lines) %>% 
  mutate(inversion.x = "Standard")
#do do reverse for major inversions


#now we'll get data for comparisions between two lines homozygous for the same inversion
inv.data = mergedata %>% 
  filter(inversion.x == inversion.y) %>%
  filter(status.x == "INV") %>% 
  filter(status.y == "INV") %>% 
  select(!c(inversion.y, status.x, status.y)) %>% 
  as.data.table(.)
inv.bind = rbind(st.data, inv.data)
```

now make density plots of different gif relatedness
```{r}
#make a seperate data set that finds the mean relatedness per method/inversion
inv.filter = inv.bind %>% 
  filter(method == "locogrm")
mean = inv.filter %>% 
  group_by(method, inversion.x, chr) %>% 
  summarise(mean.val = mean(value))
saveRDS(inv.bind, "relatedness.inversion.grm")
inv.bind2 = readRDS("relatedness.inv")
ggplot(inv.filter, aes(x = value, color = inversion.x)) + geom_density() +
  geom_vline(data=mean, aes(xintercept=mean.val, color=inversion.x),
             linetype="dashed")+
  facet_grid(chr~., scale = "free")+
  xlim(-0.075, 0.25)+
  xlab("relatedness")+
  ggtitle("Relatedness using Loco GRM")
  
```

now we'll try to make a new version of the density plots, but faceted out by the loco, instead of averaging them.

```{r}
chrom = c("2L", "2R", "3L" ,"3R", "X")
locolist = list(locogrm2l, locogrm2r, locogrm3l, locogrm3l,locogrmx)
out = foreach(f = c(1:5)) %do% {
  #f =1
  #now for loco
  locogrm = as.data.table(locolist[f])

locogrm$ids = colnames(locogrm)
locomelt = melt(locogrm, id.vars = c("ids"), variable.name = "id2")
#remove any row where id1 = id2
#and any duplicate combo of id1 and id2
locomelt$id2 = as.character(locomelt$id2)
#remove 
loco.r = subset(locomelt, ids != id2 & 
             !duplicated(cbind(pmin(ids, id2),
                               pmax(ids, id2))))
loco.r$method = "locogrm"
loco.r$chr = chrom[f]
loco.r
}
loco.out = rbindlist(out)
#now we find the median relatedness per dgrp comparison
loco.out$combo.id = paste(loco.out$ids, loco.out$id2, sep = "-")
```

